---
title: "Biomarkers and Pathways for COVID-19"
author: 
  - Ali Rahnavard
  - Chiraag Gohel
format: html
---

## Setup

1. Create a new R Project
2. Within that R Project, create a directory titled `data/`
3. Save the required data files for this vignette to that directory from [here](https://www.dropbox.com/sh/gy2ul4h36ko5rne/AADpGXEyNbYmL-0ujJzCbyUYa?dl=0)

## Tweedieverse

We first install `Tweedieverse` from GitHub and load the package.

```{r, eval = F}
devtools::install_github("himelmallick/Tweedieverse")
library(Tweedieverse)
```

We then load the metabolomics data that we wish to analyze.

```{r, eval = F}
metadata <- read.table(
  'data/metabolites_metadata.txt',
  sep = '\t',
  header = TRUE,
  fill = FALSE,
  comment.char = "" ,
  check.names = FALSE,
  row.names = 1
)

metabolites <- read.delim(
  'data/metabolites.txt',
  sep = '\t',
  header = TRUE,
  fill = T,
  comment.char = "" ,
  check.names = F,
  row.names = 1
)
```

### Running Tweedieverse

`Tweedieverse()` requires the observational data alongside the metadata, a directory path for results, and takes optional parameters for displaying results.

```{r, eval = F}
# imputation strategy
metabolites[is.na(metabolites)] <- 0 #min(metabolites, na.rm = T)/2.0

dir.create("analysis")

Tweedieverse(
  metabolites,
  metadata,
  'analysis/my_metabolites_Tweedieverse',
  max_significance = 0.1,
  plot_heatmap = T,
  plot_scatter = T,
  standardize = F
)
```

## omePath

Again, we start with installing and loading `omePath`

```{r, eval = F}
devtools::install_github('omicsEye/omePath', force = TRUE)
```

```{r, eval = F}
library(omePath)
```

omePath requires:

1. Output from a `Tweedieverse` run
2. Metadata associated with the `Tweedieverse` input
3. A mapping file that links compounds of interest to pathways

```{r, eval = F}
metabolites_Tweedieverse <- read.delim(
  "analysis/my_metabolites_Tweedieverse/all_results.tsv",
  sep = '\t',
  header = T,
  fill = F,
  comment.char = "" ,
  check.names = F,
  #row.names = NA
)

mapper_file <- read.delim(
  "data/MetaboliteStudyListwithHMDBID.csv",
  sep = ',',
  header = T,
  fill = F,
  comment.char = "" ,
  check.names = F,
  #row.names = NA
)

mapper <- read.delim(
  "data/smpdb_metabolites.tsv",
  sep = '\t',
  header = T,
  fill = F,
  comment.char = "" ,
  check.names = F,
  #row.names = NA
)
score_data_severe <- metabolites_Tweedieverse[metabolites_Tweedieverse$metadata=="Group" & metabolites_Tweedieverse$value=="Severe" ,]
rownames(score_data_severe) <- score_data_severe$feature

score_data_severe$HMDB <- mapper_file[match(score_data_severe$feature, mapper_file$Query), "HMDB"]
score_data_severe <- score_data_severe[!is.na(score_data_severe$HMDB) & score_data_severe$feature !="2-hydroxyhippurate (salicylurate)",]
rownames(score_data_severe) <- score_data_severe$HMDB
```

We then run `omePath()`, specifying various parameters.

```{r, eval = F}
omePath(
  input_data = score_data_severe,
  input_metadata = NA,
  meta <- NA,
  output = "analysis/my_deepath_enrichment_metabolite_severe_fdr1_ks",
  score_col = 'coef',
  pval_threshold = 0.05,
  fdr_threshold = NA,
  Pathway.Subject = NA,#'Metabolic',
  do_plot = TRUE,
  mapper_file = "data/smpdb_metabolites.tsv",
  method = "ks",
  min_member = 2,
  pathway_col = "Pathway",
  feature_col = "Feature")
```





